---
title: "Chloroquine Effect"
author: "Stephen R. Proulx"
date: "4/22/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(rstan)
rstan_options(auto_write = TRUE)
options(mc.cores = parallel::detectCores())

library(rethinking)
```


```{r thedata}
outcomes = tibble(treatment = c("HC","HC_AZ","None") , 
                  death = c(27,25,18) , 
                  discharge = c(70,88,140) , 
                  coded_treatment = c(1,2,3))  
```



 Here is a model where each treatment group has it's own independent mu value. This is really equivalent to running three different models, because no information is really shared between the parameters. This would not be the case if the priors were about the relationships between the parameters somehow.
```{r write_stan}
sink("model_treat_effect.stan")
cat("
    
  data {
  int <lower=1> N; // number of rows in our data matrix
  int <lower=0> death[N]; // vector of death data
  int <lower=0> discharge[N]; // vector of discharge data
  int <lower=0> coded_treatment[N];//the coded treatment level

}
parameters {
  real <lower=0,upper=1> mu[N]; // define the parameter for the probability of survival, there is one independent for each category.
}

model {
//priors
mu ~ uniform(0,1); //uniform prior

for(i in 1:N){
death[i] ~ binomial(death[i]+discharge[i] ,mu[i]);
}

}
 
    
    
    ",fill = TRUE)
sink()


```



```{r run_stan_1 , echo=FALSE}


stan_data <- c(outcomes[c("death","discharge","coded_treatment")], list(N = 3))



stanfit_chloroquine <- stan('model_treat_effect.stan', data = stan_data, chains = 4,
                      iter = 4000, seed = 2131231)
```

 
```{r print_stan_1}

print(stanfit_chloroquine ,pars=c("mu"))
pairs(stanfit_chloroquine ,pars=c("mu"))


mcmc_areas(stanfit_chloroquine,area_method = "equal height",prob=.95,pars=c("mu[1]","mu[2]","mu[3]"))

```


TO make things a little simpler and make it easier to ask our question we can combine the two categories with HC (i.e. HC and HC+AZ)
```{r run_stan_2 , echo=FALSE}


outcomes = tibble(treatment = c("HC","None") , 
                  death = c(27+25,18) , 
                  discharge = c(70+88,140) , 
                  coded_treatment = c(1,2))  


stan_data <- c(outcomes[c("death","discharge","coded_treatment")], list(N = 2))



stanfit_chloroquine2 <- stan('model_treat_effect.stan', data = stan_data, chains = 4,
                      iter = 4000, seed = 2131231)
```


```{r print_stan_2 }

print(stanfit_chloroquine2 ,pars=c("mu"))
pairs(stanfit_chloroquine2 ,pars=c("mu"))


mcmc_areas(stanfit_chloroquine2,area_method = "equal height",prob=.95,pars=c("mu[1]","mu[2]"))

```



Here we do the work to see if the two parameters are really different. We do this by getting just the list of sampled parameters using the extract function, and then looking at the difference between the two parameters for each sampled parameter set. Becasue there is no correlation among mu's, this really doesn't change our interpretation.
```{r diff_plot }

 post=extract(stanfit_chloroquine2) %>% as_tibble() %>% select(- "lp__") %>% mutate(diff = mu[2]-mu[1]) 

post=as.data.frame(stanfit_chloroquine2) %>% as_tibble() %>% select(- "lp__")%>% 
  rename(mu1="mu[1]",mu2="mu[2]") %>% 
  mutate(diff = mu2-mu1) 



mcmc_areas(post, area_method = "equal height",  prob=.95)

```




## Lumping the HC together and modeling the difference in death rate between no-treatment and HC
This one assumes there are only two treatment types, lumping the HC and HC+AZ together. It also models the no treatment and HC group death rates as an additive difference on the global mean, on the logit scale. This means that we have a parameter, mu, which is on the normal probability scale (i.e. between 0 and 1) and a treatment effect on the logit scale, so between -Infinity and Infinity. This works by transforming mu with the logit function: logit(mu), then adding the displacement parameter, rho, then inverse transforming back to the probability scale with inv_logit.
```{r write_stan_2}
sink("model_treat_effect_diff2.stan")
cat("
    
  data {
  int <lower=1> N; // number of rows in our data matrix
  int <lower=0> death[N]; // vector of death data
  int <lower=0> discharge[N]; // vector of discharge data
  int <lower=0> coded_treatment[N];//the coded treatment level
  real <lower=0> rho_prior; // sd for rho

  }
  transformed data{
  real indicator_treatment[N];
  for(i in 1:N){
  indicator_treatment[i] = (coded_treatment[i] == 0 ) * (-1) + (coded_treatment[i] == 1)*1;
  }
  }
parameters {
  real <lower=0,upper=1> mu; // define the parameter for the probability of survival overall
  real rho; // this parameter will be the difference between the groups and the overall mean. It is over-parameterized, but that's ok!
}

model {
//priors
mu ~ uniform(0,1); //uniform prior
rho ~ normal(0,rho_prior); // Strong prior that there is not a difference between treatments. 

for(i in 1:N){
death[i] ~ binomial(death[i]+discharge[i] ,inv_logit( logit(mu) + rho *indicator_treatment[i] ) );
}

}
 
    
    
    ",fill = TRUE)
sink()


```


```{r}

stan_data <- c(outcomes[c("death","discharge","coded_treatment")], list(N = 2 , rho_prior=.5))



stanfit_chloroquine_diff <- stan('model_treat_effect_diff2.stan', data = stan_data, chains = 8,
                      iter = 10000, seed = 2131231)
```
```{r print_stan_2 }

print(stanfit_chloroquine_diff ,pars=c("mu","rho"))
pairs(stanfit_chloroquine_diff ,pars=c("mu","rho"))


mcmc_areas(stanfit_chloroquine_diff,area_method = "equal height",prob=.95,pars=c("mu","rho"))

```
