---
title: "ChimpTest"
author: "Stephen R. Proulx"
date: "12/31/2020"
output: html_document
---
 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(rstan)
rstan_options(auto_write = TRUE)
options(mc.cores = parallel::detectCores())


library(bayesplot)
library(rethinking)
```


```{r}
## R code 11.1
library(rethinking)
data(chimpanzees)
d <- chimpanzees
```

Most chimps pull the right lever
```{r}
## R code 11.2
d<- mutate(d , treatment = 1 + prosoc_left + 2*condition)

#d$treatment <- 1 + d$prosoc_left + 2*d$condition
 
## R code 11.4
m11.1 <- quap(
    alist(
        pulled_left ~ dbinom( 1 , p ) ,
        logit(p) <- a ,
        a ~ dnorm( 0 , 1.5 )
    ) , data=d )

```

How's that prior looking, pretty crap until you turn the sd down to 1.5!
```{r}
## R code 11.5
set.seed(1999)
prior <- extract.prior( m11.1 , n=1e6 )

## R code 11.6
p <- inv_logit( prior$a )
dens( p , adj=0.1 )
```

Check the prior for the model with treatment effects. Note that the structure of the extracted prior is not a dataframe, so I've converted it to a tibble to make mutate commands easier. 
```{r}
## R code 11.7
m11.2 <- quap(
    alist(
        pulled_left ~ dbinom( 1 , p ) ,
        logit(p) <- a + b[treatment] ,
        a ~ dnorm( 0 , 1.5 ),
        b[treatment] ~ dnorm( 0 , 1)
    ) , data=d )

set.seed(1999)
prior <- extract.prior( m11.2 , n=1e6 ) %>% as_tibble()
p <- prior %>% mutate(delta_prop = abs(inv_logit(a+ b[,1] ) - inv_logit(a+ b[,4])))
  
## R code 11.8
dens( p$delta_prop , adj=0.1 )
```


Now we'll try our reasonable priors and see that the difference between treatments is about 0.1 on the probability scale (i.e. 0 to 1)
```{r}


## R code 11.9
m11.3 <- quap(
    alist(
        pulled_left ~ dbinom( 1 , p ) ,
        logit(p) <- a + b[treatment] ,
        a ~ dnorm( 0 , 1.5 ),
        b[treatment] ~ dnorm( 0 , 0.5 )
    ) , data=d )
set.seed(1999)
prior <- extract.prior( m11.3 , n=1e4 )%>%as_tibble()

p <- prior %>% mutate(delta_prop = abs(inv_logit(a+ b[,1] ) - inv_logit(a+ b[,4])))
mean( p$delta_prop )
  
```


```{r}
## R code 11.10
# trimmed data list
dat_list <- list(
    pulled_left = d$pulled_left,
    actor = d$actor,
    treatment = as.integer(d$treatment) )

## R code 11.11
m11.4 <- ulam(
    alist(
        pulled_left ~ dbinom( 1 , p ) ,
        logit(p) <- a[actor] + b[treatment] ,
        a[actor] ~ dnorm( 0 , 1.5 ),
        b[treatment] ~ dnorm( 0 , 0.5 )
    ) , data=dat_list , chains=4 , log_lik=TRUE )
precis( m11.4 , depth=2 )
```


```{r}
## R code 11.12
post <- extract.samples(m11.4)  
p_left <- inv_logit( post$a )
precis( as.data.frame(p_left) )
```


```{r}
## R code 11.13
labs <- c("R/N","L/N","R/P","L/P")
plot( precis( m11.4 , depth=2 , pars="b" ) , labels=labs )
precistab <- as.data.frame(precis( m11.4 , depth=2 , pars="b" ) ) %>% 
  mutate(treatment = row_number())%>%
  rename(lower = "5.5%" , upper = "94.5%")
ggplot(precistab,aes(x=treatment, y= mean))+
  geom_point()+
  geom_errorbar(aes(ymin = lower, ymax=upper))
```


```{r}
## R code 11.14
 
post2<-as_tibble(post)

diffs <- select(post2, b) %>% mutate(db13 = b[,1]-b[,3], db24 = b[,2]-b[,4])

mcmc_areas(diffs, pars=c("db13","db24") , prob = 0.87)
  
```

I don't know how to draw this on my own...
```{r}
## R code 11.15
pl <- by( d$pulled_left , list( d$actor , d$treatment ) , mean )
pl[1,]

## R code 11.16
plot( NULL , xlim=c(1,28) , ylim=c(0,1) , xlab="" ,
    ylab="proportion left lever" , xaxt="n" , yaxt="n" )
axis( 2 , at=c(0,0.5,1) , labels=c(0,0.5,1) )
abline( h=0.5 , lty=2 )
for ( j in 1:7 ) abline( v=(j-1)*4+4.5 , lwd=0.5 )
for ( j in 1:7 ) text( (j-1)*4+2.5 , 1.1 , concat("actor ",j) , xpd=TRUE )
for ( j in (1:7)[-2] ) {
    lines( (j-1)*4+c(1,3) , pl[j,c(1,3)] , lwd=2 , col=rangi2 )
    lines( (j-1)*4+c(2,4) , pl[j,c(2,4)] , lwd=2 , col=rangi2 )
}
points( 1:28 , t(pl) , pch=16 , col="white" , cex=1.7 )
points( 1:28 , t(pl) , pch=c(1,1,16,16) , col=rangi2 , lwd=2 )
yoff <- 0.01
text( 1 , pl[1,1]-yoff , "R/N" , pos=1 , cex=0.8 )
text( 2 , pl[1,2]+yoff , "L/N" , pos=3 , cex=0.8 )
text( 3 , pl[1,3]-yoff , "R/P" , pos=1 , cex=0.8 )
text( 4 , pl[1,4]+yoff , "L/P" , pos=3 , cex=0.8 )
mtext( "observed proportions\n" )
```


```{r}
## R code 11.17
dat <- list( actor=rep(1:7,each=4) , treatment=rep(1:4,times=7) )
p_post <- link( m11.4 , data=dat ) %>% as_tibble()
post_log <- p_post %>% gather(  "chimp_treatment" ,  "prop_left")

p_mu <- apply( p_post , 2 , mean )
p_ci <- apply( p_post , 2 , PI )
```


```{r}
## R code 11.18
d$side <- d$prosoc_left + 1 # right 1, left 2
d$cond <- d$condition + 1 # no partner 1, partner 2

## R code 11.19
dat_list2 <- list(
    pulled_left = d$pulled_left,
    actor = d$actor,
    side = d$side,
    cond = d$cond )
m11.5 <- ulam(
    alist(
        pulled_left ~ dbinom( 1 , p ) ,
        logit(p) <- a[actor] + bs[side] + bc[cond] ,
        a[actor] ~ dnorm( 0 , 1.5 ),
        bs[side] ~ dnorm( 0 , 0.5 ),
        bc[cond] ~ dnorm( 0 , 0.5 )
    ) , data=dat_list2 , chains=4 , log_lik=TRUE )

## R code 11.20
compare( m11.5 , m11.4 , func=PSIS )
```


```{r}
## R code 11.21
post <- extract.samples( m11.4 , clean=FALSE )
str(post)

## R code 11.22
m11.4_stan_code <- stancode(m11.4)
m11.4_stan <- stan( model_code=m11.4_stan_code , data=dat_list , chains=4 )
compare( m11.4_stan , m11.4 )

## R code 11.23
post <- extract.samples(m11.4)
mean( exp(post$b[,4]-post$b[,2]) )

## R code 11.24
data(chimpanzees)
d <- chimpanzees
d$treatment <- 1 + d$prosoc_left + 2*d$condition
d$side <- d$prosoc_left + 1 # right 1, left 2
d$cond <- d$condition + 1 # no partner 1, partner 2
d_aggregated <- aggregate(
    d$pulled_left ,
    list( treatment=d$treatment , actor=d$actor ,
          side=d$side , cond=d$cond ) ,
    sum )
colnames(d_aggregated)[5] <- "left_pulls"

## R code 11.25
dat <- with( d_aggregated , list(
    left_pulls = left_pulls,
    treatment = treatment,
    actor = actor,
    side = side,
    cond = cond ) )

m11.6 <- ulam(
    alist(
        left_pulls ~ dbinom( 18 , p ) ,
        logit(p) <- a[actor] + b[treatment] ,
        a[actor] ~ dnorm( 0 , 1.5 ) ,
        b[treatment] ~ dnorm( 0 , 0.5 )
    ) , data=dat , chains=4 , log_lik=TRUE )

## R code 11.26
compare( m11.6 , m11.4 , func=PSIS )

## R code 11.27
# deviance of aggregated 6-in-9
-2*dbinom(6,9,0.2,log=TRUE)
# deviance of dis-aggregated
-2*sum(dbern(c(1,1,1,1,1,1,0,0,0),0.2,log=TRUE))
```


