---
title: "Lab 4"
author: "Stephen R. Proulx"
date: "12/27/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(rethinking)
```
# Bayesian Statistical Modeling Winter 2021
# Lab Exercise, Week 4
# 1/28/2021

*When is this lab due?* Labs are due on the Thursday after they are assigned. However, in many cases you can complete them during the lab period itself.   This assignment is due on Thursday, 2/4/2021. Submit homework to gradescope: *Link here*

## Multiple Regression 
Fox dataset 

```{r}
data("foxes")  
d=as_tibble(foxes) 

d$F <- standardize(d$avgfood)
d$W <- standardize(d$weight)
d$A <- standardize(d$area)
d$G <- standardize(d$groupsize)
```

```{r}
ggplot(data=d,aes(x=avgfood,y=weight,group=G,color=G)) +geom_point()
```


```{r}
m3 <- quap(
alist(
W ~ dnorm( mu , sigma ),
mu <- a + bF*F + bG*G,
a ~ dnorm(0,0.2),
c(bF,bG) ~ dnorm(0,0.5),
sigma ~ dexp(1)
), data=d )
precis(m3)
```


compare actual weights with predicted weights from the model?