---
title: 'Day 7: Multicollinearity'
author: "Stephen R. Proulx"
date: "1/10/2021"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(rethinking)
source("../Resources/helper.R")

```

# Today's objectives:  

* Perform mutliple regression on simulated data to explore how regression models do and do not capture causal processes.  
* Use posterior simiulation data to see how models where no slope is different from 0 can still predict data accurately  
* Conduct multiple regression with a real dataset that exhibits multicollinearity   

 


## Effects of Multicollinearity

Here we start with simulated data where we assume that legs are 40-50% of total height and that there is some difference between measured right and left leg length.
```{r}
n <- 100
set.seed(5)

d <- 
  tibble(index = seq(1:n) , height    = rnorm(n, mean = 10, sd = 2),
         leg_prop  = runif(n, min = 0.4, max = 0.5)) %>% 
  mutate(leg_left  = leg_prop * height + rnorm(n, mean = 0, sd = 0.02),
         leg_right = leg_prop * height + rnorm(n, mean = 0, sd = 0.02))
```


Left and right leg lengths are highly correlated:
```{r}
ggplot(d,aes(x = leg_left, y = leg_right)) +
  geom_point(alpha = 1/2 ) 
```

And so is either leg length with total height, although in our model there is a bit more scatter.
```{r}
ggplot(d,aes(x = leg_left, y = height)) +
  geom_point(alpha = 1/2 )  
ggplot(d,aes(x = leg_right, y = height)) +
  geom_point(alpha = 1/2 )  
```


### Is leg length a good predictor of total height?
Go through the process of fitting models of these data. Create one model that does not use either leg, one with only the left leg, one with only the right leg, and a model that has both legs. Give each model a distinctive name so you can use the output again later.   

Use precis and whichever methods of plotting posterior distributions of parameters you like to get an idea of what is happening in each of these models. 


### Posterior predictions for leg length simulation
Now create plots of the expected data that would come out of each of the fitted models. Try and get something that has the actual height of the individual on the x-axis and shows the mean and range of likely height values predicted for that individual. You may also want to include the actual height on the graph as well, or another way of drawing a line with slope 1. 



### Collinearity in real data: energy content of milk
Here is a dataset that measures the calorie content of mammal milk as well as the fat and lactose content of the milk.

```{r}
data(milk)
d <- milk %>% as_tibble() %>% 
  rowid_to_column("index") %>%
  mutate(K=standardize(kcal.per.g),
         F=standardize(perc.fat),
         L=standardize(perc.lactose))

ggplot(data=d , aes(y=K, x= F))+geom_point()

ggplot(data=d , aes(y=K, x= L))+geom_point()


ggplot(data=d , aes(y=F, x= L))+geom_point()
```


Here is the quap call to conduct multiple regression with F and L as predictors.

Use this model fit to explore how including the predictor affects our posterior simulations when we drop F and/or L from the predictions by setting it to the mean value in the popultion.


```{r}
m6.5.full <- quap(
    alist(
        K ~ dnorm( mu , sigma ) ,
        mu <- a + bF*F + bL*L ,
        a ~ dnorm( 0 , 0.2 ) ,
        bF ~ dnorm( 0 , 0.5 ) ,
        bL ~ dnorm( 0 , 0.5 ) ,
        sigma ~ dexp( 1 )
    ) ,
    data=d )
precis( m6.5 )

```

