---
title: "Final Assessment"
author: "Stephen R. Proulx"
date: "3/10/2021"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(rethinking)
source("../Resources/helper.R")
```


## Short Awnswer

### Question 1

Consider this dataset where $y$ is the number of eggs laid by individual fruit flies who are put into vials with $x$ microliters of yeast paste. 
```{r}
data <- tibble( y = c(35,43,54,33,65,54,72,55,66), x = c(5,5,5,10,10,10,15,15,15))
```

Here is a model specification in ulam.
```{r}
model <- ulam(alist(
  y ~ dpois(lambda),
  lambda <- a + b * x,
  a ~ dnorm(0,3),
  x ~ dnorm(0,1)
) ,  data=data)
```

#### What's wrong with this ulam model specification?

### Question 2
 The dataframe has a column $y$ that are body lenght measurements for a species of lizard. Column $x_1$ is 0 for male lizards and 1 for female lizards. Column $x_2$ is 0 for brown throated lizards and 1 for green throated lizards. We add another 4 columns that are indicator variables for each of the 4 states that a lizard can be in (male/female X brown/green), defining $x_{i,j}=1$ if $x_1=i$ and $x_2=j$. 
 
Below are three model definitions (partial definitions, I have not written out the priors).

Model 1
$$
 y \sim \mathrm{normal}(\mu,\sigma) \\
 \mu = a + b_1 x_1 + b_2 x_2
 
$$
Model 2
$$
 y \sim \mathrm{normal}(\mu,\sigma) \\
 \mu = a + b_1 x_1 + b_2 x_2 + b_{12} x_1 x_2
 
$$
Model 3
$$
 y \sim \mathrm{normal}(\mu,\sigma) \\
 \mu =  b_{0,0} x_{0,0} + b_{0,1} x_{0,1} + b_{1,0} x_{1,0} + b_{1,1} x_{1,1}
$$
#### Which of these models can be considered models with interaction? Explain why.


#### Consider these values for the means of categories below: 

sex   color    mean length    
----- -------  --------------
M      B         13
M      G         15
F      B         16
F      G         20

Can you choose values of the parameters for each of the three models that would have the mean, $\mu$, be equal to these mean values in each category? 


### question 3
Consider this dataset:
```{r}
data <- tibble( x=seq(1,10),y=c(4.4,6.0,4.2,6.9,5.4,3.3,1.1,2.7,1.3,5.5))

ggplot(data,aes(x=x,y=y))+ 
  geom_point()+
  ylim(0,10)
```

Here are two potential models to fit these data:
```{r}
m.1 <- ulam(alist(
  y ~ dnorm(mu, sigma),
  mu <- a + b1*x,
  a ~ dnorm(5,2.5),
  b1 ~ dnorm(0,1),
  sigma ~ dexp(1)
))
```

```{r}
m.2 <- ulam(alist(
  y ~ dnorm(mu, sigma),
  mu <- a + b1*x + b2*x^2 + b3*x^3,
  a ~ dnorm(5,2.5),
  b1 ~ dnorm(0,1),
  b2 ~ dnorm(0,1),
  b3 ~ dnorm(0,1),
  sigma ~ dexp(1)
))
```

#### After fitting the models you can compute the probability of observing the data given the parameters. Which model do you expect to fit the data best, and why?





####  Now consider this additional point is discovered and added to the dataset. Which model do you expect will fit this dataset better and why?
```{r}
data <- tibble( x=seq(1,10),y=c(4.4,6.0,4.2,6.9,5.4,3.3,1.1,2.7,1.3,5.5))
new_data <- tibble( x=c(7), y= c(5.0) )

ggplot(data,aes(x=x,y=y))+ 
  geom_point()+
  geom_point(data=new_data,color="red")+
  ylim(0,10)
```


## Analaysis Challenge
These data are part of a study conducted by Kelly Thomasson in my lab that involved evolving yeast for that were being passaged through fruit flies. The yeast were divided into replicate populations that evolved under "control" conditions (no flies), or under "treatment" conditions (with flies). After evolution, she performed an assay to see if the production of spore cells relative to vegetative cells had evolved. To do this, she counted the number of cells that were spores and the number that were vegetative. 

In the dataset I am providing, there are two strains of yeast, "AM" from an american oak forrest, and "WE", a european wine production strain. Both of these strains were assayed for sporulation before evolution, this is the "ancestral" treatment. For the ancestral treatment, they were assayed 12 times each.

In the evolution treatments, each strain was split into 4 populations that evolved independently, and then each strain was assayed 4 times. In the dataset, "unique_replicate_id" uniquely identifies each observation of a group of cells. 

Come up with an approach to analyzing these data, with an eye towards determining what affect the evolution treatment has and whether or not the strains differ in the way they evolved. You may either use a model selection method based on WAIC or an approach that measures contrasts between parameters to do this.  Discuss the results both on the logit scale and on the probability scale. 


```{r}

load("sporedata.RDS")

spore_data <- spore_data %>%
  mutate(strain_indicator = (strain=="AM")*0 + (strain=="WE")*1,
         strain_index = (strain=="AM")*1 + (strain=="WE")*2,
         treat_index = (treatment.type == "ancestral")*1+(treatment.type == "control")*2+(treatment.type == "treatment")*3,
         total=spore_count+veg_count)
       

spore_data  
```



