---
title: "Lab 9"
author: "Stephen R. Proulx"
date: "12/27/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(rethinking)
library(bayesplot)
source("../Resources/helper.R")
```
# Bayesian Statistical Modeling Winter 2021
# Lab Exercise, Week 9
# 3/4/2021

*When is this lab due?* Labs are due on the Thursday after they are assigned. However, in many cases you can complete them during the lab period itself.   This assignment is due on Thursday, 3/11/2021.  




# Reedfrog survival: Models with different predictors
In class we worked with some multilevel models of the reedfrog data where we either allowed each tank to be fit as a "random" effect, or we had a random effect and an effect of predation that was dependent on density. Technically, this is already an interaction model. 

Today you will make models that include two of the predictor variables from the dataset, namely tadpole size and predation. 

Consider a model with no predictor (i.e. just the random tank effect), with predation, with size, with size and predation, and with the interaction between size and predation.

Here's the data- you will need to add some columns to correctly format the predictors for your model. 
```{r}
library(rethinking) 
data(reedfrogs)
d <- reedfrogs
```

## Build and sample the models
Write code to build and sample the 5 models. Run 4 chains for 3000 iterations. Inspect the `precis` to ensure that the chains have converged. 



## Compare the inferences on sigma between the models
How do the models differ in their posterior distributions for sigma? Why might this be?



## Compute and compare WAIC.
Interpret the WAIC values, their differences and SE values, as well as the WAIC penalty terms.




## Compute contrasts for size and for the effect of size on predation (i.e. the interaction terms)
Using the model with interactions, compute from the posterior, the difference between the effect of small and large tadpoles. This is the "size effect contrast". 

Also compute the difference between the effect of predation for small and for large tadpoles. This is the contrast for the size-predation interaction term.

What can you conclude from these contrasts?






