---
title: "Lab 6"
author: "Stephen R. Proulx"
date: "12/27/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(rethinking)
library(bayesplot)
source("../Resources/helper.R")
```
# Bayesian Statistical Modeling Winter 2022
# Lab Exercise, Week 7
# 2/17/2022

*When is this lab due?* Labs are due on the Thursday after they are assigned. However, in many cases you can complete them during the lab period itself.   This assignment is due on Thursday, 2/24/2022. Submit homework to gradescope.

## Wine tasting 
We'll be using data on wine ratings from professional wine judges. In this dataset wines are scored by judges in multiple flights. 

Load the data and standardize the score so that it's mean is 0. We will also make columns that are numeric representations of the judge name and wine, and an indicator variable column for flight.
```{r}
data(Wines2012)
d <- Wines2012 %>% 
  as_tibble() %>%
  mutate(S = standardize(score),
         fid=as.integer(flight) -1) 
```

Have a look at the data. As potential predictors, we will be using the nationality of the judge, `judge.amer`, the nationality of the wine, `wine.amer`, and whether the wine is white or red, `fid`. Each of these is binary and in the dataframe is either 0 or 1, which can make it easier to turn parameters on and off by simply multiplying by the value of the predictor. 

### An additive model for wine scores
First create a model where each of the three predictors has an additive effect on score. Assume that scores are normally distributed and that the standard deviation is the same for all categories. 

You can visualize the results using something like this:
```{r}
samples <- extract.samples(your.model) %>%
  select(-sigma)

bayesplot::mcmc_intervals(samples)
```




### 2 way interactions
Now include interaction terms and visualize the estimates. Compare the models in terms of their WAIC. What can you conclude from the WAIC scores?


### All the interactions
There are two straightforward ways to do this. One is to add onto your last model by including the 3-way interaciton effect. The other is to write a sort of binary model that has a parameter for each possible state of the 3 binary predictor variables. 

Again compare the WAIC scores. 


